Modelling last glacial cycle ice dynamics in the Alps

### 2018-08-17 articles/cycle/alps-cycle.tex
Update reference to Cohen et al. (2018).

    The spatial extent and thickness of Alpine glaciers during the LGM, an
    integrated footprint of thousands of years of climate history and glacier
    dynamics, have been reconstructed from moraines and trimlines across the
    mountain range \citep[e.g.,][]{Penck.Bruckner.1909, Castiglioni.1940,
    Husen.1987, Bini.etal.2009, Coutterand.2010}. The assumption that
    trimlines, the upper limit of glacial
    erosion, represent the maximum ice surface elevation, has
    been repeatedly invalidated in other glaciated regions of the globe
    \citep[e.g.,][]{Kleman.1994, Kleman.etal.2010, Fabel.etal.2012,
    Ballantyne.Stone.2015}. Although
    no evidence for cold-based glaciation has been reported in the Alps to
    date, it is supported by numerical modelling \citep{Blatter.Haeberli.1984,
    Haeberli.Schluchter.1987, Cohen.etal.2017}. Alpine ice flow
    patterns were primarily controlled by subglacial topography, but there is
    evidence for flow accross major mountain passes
    \citep[e.g.,][]{Coutterand.2010, Kelly.etal.2004, Husen.2011}, and
    self-sustained ice domes \citep{Bini.etal.2009}. Finally, the timing of the
    LGM in the Alps \citep{Ivy-Ochs.etal.2008, Monegato.etal.2017} is in good
    agreement with the maximum expansion of continental ice sheets recorded by
    the Marine Oxygen Isotope Stage (MIS) 2
    \citep[29--14\,ka;][]{Lisiecki.Raymo.2005}. Regional variation between
    different piedmont
    lobes exists \citep[Fig.~5]{Wirsig.etal.2016}, but it is unclear whether
    this relates to climate or glacier dynamics \citep{Monegato.etal.2017},
    uncertainties in the dating methods, or both.

@@

    In the upper Rhone Valley, the maximum ice surface elevation reached
    during MIS~2 in our simulation, corrected for bedrock deformation, is
    consistently modelled to have occured several hundred metres above the
    observed trimline elevations (Fig.~\ref{fig:trimlines}a and b), with a mean
    difference of 861\,m (and a standard deviation 197\,m,
    Fig.~\ref{fig:trimlines}b), which is significant in respect to the modelled
    maximum ice thickness in the Rhone Valley (and the Alps) of 2586\,m. This
    result depends on uncertain basal sliding and ice rheological parameters
    to which the model sensitivity was not tested here.  However, regional
    model sensitivity tests show that a modelled surface compatible with the
    trimline elevations is incompatible with the documented overspilling flow
    across the Simplon Pass \citep{Becker.2017, Becker.etal.2017}, and LGM
    palaeo-climate proxy records \citep{Cohen.etal.2017}. Unfortunately,
    validation through the bedrock uplift rate \citep[cf.][]{Kuchar.etal.2012}
    is not doable in the Alps due its lower values, active tectonics and
    uncertainties on geological properties \citep[cf.][]{Mey.etal.2016}.

@@

    In the upper Rhone Valley, observed trimlines are often located near the
    LGM cold-temperate basal thermal transition or within cold-based areas
    (Fig.~\ref{fig:trimlines}c). The remaining discrepancies between the
    trimlines locations and the modelled basal thermal boundary may relate to
    temporal migrations of the basal thermal boundary, an absence of sliding in
    warm-based areas, and levelling of small-scale topographic features in the
    1\,km horizontal grid. They call for more detailed comparisons
    spanning the entire Alpine range and specific sensitivity studies to
    relevant basal sliding and ice rheological parameters, and to the uncertain
    subglacial topography. However, the presence of an upper layer of
    cold ice during the LGM, already found at high altitude in the much warmer
    climate of today \citep[e.g.,][]{Suter.etal.2001, Bohleber.etal.2017}, is
    inevitable \citep{Blatter.Haeberli.1984, Haeberli.Schluchter.1987,
    Cohen.etal.2017}. The bedrock themal model yield significant volumes of
    frozen ground periglacially north of the Alps and subglacially under the
    highest mountains, which is also in agreement with previous studies
    \citep{Haeberli.etal.1984, Lindgren.etal.2016, Cohen.etal.2017}.

>>>>

    The spatial extent and thickness of Alpine glaciers during the LGM, an
    integrated footprint of thousands of years of climate history and glacier
    dynamics, have been reconstructed from moraines and trimlines across the
    mountain range \citep[e.g.,][]{Penck.Bruckner.1909, Castiglioni.1940,
    Husen.1987, Bini.etal.2009, Coutterand.2010}. The assumption that
    trimlines, the upper limit of glacial
    erosion, represent the maximum ice surface elevation, has
    been repeatedly invalidated in other glaciated regions of the globe
    \citep[e.g.,][]{Kleman.1994, Kleman.etal.2010, Fabel.etal.2012,
    Ballantyne.Stone.2015}. Although
    no evidence for cold-based glaciation has been reported in the Alps to
    date, it is supported by numerical modelling \citep{Blatter.Haeberli.1984,
    Haeberli.Schluchter.1987, Cohen.etal.2018}. Alpine ice flow
    patterns were primarily controlled by subglacial topography, but there is
    evidence for flow accross major mountain passes
    \citep[e.g.,][]{Coutterand.2010, Kelly.etal.2004, Husen.2011}, and
    self-sustained ice domes \citep{Bini.etal.2009}. Finally, the timing of the
    LGM in the Alps \citep{Ivy-Ochs.etal.2008, Monegato.etal.2017} is in good
    agreement with the maximum expansion of continental ice sheets recorded by
    the Marine Oxygen Isotope Stage (MIS) 2
    \citep[29--14\,ka;][]{Lisiecki.Raymo.2005}. Regional variation between
    different piedmont
    lobes exists \citep[Fig.~5]{Wirsig.etal.2016}, but it is unclear whether
    this relates to climate or glacier dynamics \citep{Monegato.etal.2017},
    uncertainties in the dating methods, or both.

@@

    In the upper Rhone Valley, the maximum ice surface elevation reached
    during MIS~2 in our simulation, corrected for bedrock deformation, is
    consistently modelled to have occured several hundred metres above the
    observed trimline elevations (Fig.~\ref{fig:trimlines}a and b), with a mean
    difference of 861\,m (and a standard deviation 197\,m,
    Fig.~\ref{fig:trimlines}b), which is significant in respect to the modelled
    maximum ice thickness in the Rhone Valley (and the Alps) of 2586\,m. This
    result depends on uncertain basal sliding and ice rheological parameters
    to which the model sensitivity was not tested here.  However, regional
    model sensitivity tests show that a modelled surface compatible with the
    trimline elevations is incompatible with the documented overspilling flow
    across the Simplon Pass \citep{Becker.2017, Becker.etal.2017}, and LGM
    palaeo-climate proxy records \citep{Cohen.etal.2018}. Unfortunately,
    validation through the bedrock uplift rate \citep[cf.][]{Kuchar.etal.2012}
    is not doable in the Alps due its lower values, active tectonics and
    uncertainties on geological properties \citep[cf.][]{Mey.etal.2016}.

@@

    In the upper Rhone Valley, observed trimlines are often located near the
    LGM cold-temperate basal thermal transition or within cold-based areas
    (Fig.~\ref{fig:trimlines}c). The remaining discrepancies between the
    trimlines locations and the modelled basal thermal boundary may relate to
    temporal migrations of the basal thermal boundary, an absence of sliding in
    warm-based areas, and levelling of small-scale topographic features in the
    1\,km horizontal grid. They call for more detailed comparisons
    spanning the entire Alpine range and specific sensitivity studies to
    relevant basal sliding and ice rheological parameters, and to the uncertain
    subglacial topography. However, the presence of an upper layer of
    cold ice during the LGM, already found at high altitude in the much warmer
    climate of today \citep[e.g.,][]{Suter.etal.2001, Bohleber.etal.2017}, is
    inevitable \citep{Blatter.Haeberli.1984, Haeberli.Schluchter.1987,
    Cohen.etal.2018}. The bedrock themal model yield significant volumes of
    frozen ground periglacially north of the Alps and subglacially under the
    highest mountains, which is also in agreement with previous studies
    \citep{Haeberli.etal.1984, Lindgren.etal.2016, Cohen.etal.2018}.
